Synthesis of Linear Functional Programs

### 2021-07-14 article/article.tex
fixes according to prof

@@

\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita - Nº XXXXX \\
    Supervisor: Bernardo Toninho}
\date{July 2021}
\institute{NOVA School of Science and Technology}

@@

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program (i.e.~generating code) from a high-level specification.
%
Synthesis can be seen as a means to improve a programmer's productivity and/or
program correctness (i.e. through suggestion and/or autocompletion), or as a
tool to automate certain parts of the programming process, to name a few
examples.

Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind of
specification, program synthesizers have to deal with two main inherent sources
of complexity: searching over a vast space of (potentially) valid programs, and
interpreting user intent.
%
In this work we'll explore type-based synthesis where specifications take the
form of linear types, which limit resource usage in programs, in combination
with polymorphism.  Concretely, we will explore the proof-theoretic foundations
of linear types via the Curry-Howard correspondence with linear logic, viewing
program synthesis under the lens of (linear logic) bottom-up proof search.

@@

At the core of the synthesis process, and of the developed synthetiser \synname,
is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
resource management. It connects the concepts seen in
background~\ref{sec:background} -- a type system
consisting of inference rules in sequent calculus for linear logic connectives;
with resource management, and focusing, as formalized techniques to,
respectively, solve resource related non-determinism, and to eliminate from the
proof search don't-care non-determinism.  Guided solely by this system's formal
representation\ref{sec:formal_system}, we can (and did) create a very robust
synthetiser for a select part of the linear types that's able to synth
programs successfully from e.g. the major linear logic equivalences. For example,
given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
the program $\lambda a \rightarrow \lambda b \rightarrow$ let $c \tensor d$ =
$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.

@@

However, a programming language only with the canonical linear types lacks
expressiveness. To enhance \synname\ and its language, synth more interesting
programs -- and thus empirically proving the feasability of more relevant
synthesis in a linear context, we extend the syntax and type system with
recursiveness and richer
types: -- polimorfic, refinements, adts parametrizáveis...  todo

\section{Synthesis with Linear Types, Polymorphism and Refinements}\label{sec:formal_system}
\todo[inline]{Encurtar titulo, mas algo nesta onda que descreve
  ``tudo''}

@@

With this section we aim to formalize the rules that guide
synthesis from more expressive specifications, alongside the already
well defined rules that model the core of the synthesis process, resulting in
a \emph{sound} set of rules that hopes to serve as a source of ideas
\todo{ nao me parece\dots As regras caracterizam o que esta
  implementado, e esse o objectivo, nao e ser uma fonte de ideias\dots}
  \todo{R: não parece ser sound ?}
on taming synthesis of recursive functions with richer types in a
linear context, as well as describe the system in enough detail for
the synthetiser to be reproducible by a theory-driven implementation.

Even so, note that \emph{soundness} and \emph{completeness} have two very
different meanings, and our \emph{sound} set of rules only guarantees we won't
synth incorrect programs, but still hides a possibly infinite number of correct
proofs; and that the valid programs derivable through it reflect the somewhat
subjective trade-offs we decided on. Different choices and approaches outside
the core might lead to completely distinct synths and spaces of valid programs.

\mypara{Core Rules} The core system comprises of the sequent calculus rules for
linear logic connectives with resource-management and focusing. As
explained in\ref{sec:background} \todo{se o background nao vai ser
  incluido no documento principal, nao deves referenciar assim tao
  directamente para o conteudo principal.}, focusing separates the proof search into two
phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later try to
invert, moving them to the linear context ($\Delta$) when we fail to.

@@

Putting together linear logic, the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synth a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synth a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x \rightarrow M$ and the output linear context is
$\Delta'$.

\todo{a sintaxe $\lambda x \rightarrow M$ nao e mt standard, usa
  $\lambda x . M$}

@@

    % -o R
    \infer*[right=($\lolli R$)]
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
    \notin \Delta'}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x \rightarrow M : A
    \lolli B \Uparrow}

@@

The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a named sum type of any type, i.e. a named type that can be instantiated by one of many
tags (also known as constructors) that take exactly one fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which might be constructed with zero, one, or two
propositions of type $A$, using its syntax: $\textrm{data Container} =
\textrm{None}\ 1\ \vert\ \textrm{One}\ A\ \vert\ \textrm{Two}\ (A * A)$.

@@

\todo[inline]{Não uses Foo e Bar. $T_1$, $T_2$ , $\dots$ para
  tipos. $C_1$, $C_2$, $\dots$ para os construtores.}

@@

where the ADT $T$ and its constructors stand for any ADT with form
\[
    \textrm{data}\ T = C_1\ X_1\ \vert\ C_2\ X_2\ \vert\
\dots\ \vert\ C_n\ X_n
\]

@@

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning the type a constructor needs might
be the type the constructor constructs \todo{ Na escrita, repetition does not
legitimize. Constructors can take as arguments values of the type they are
defining, ou algo assim}. Seemingly small, this change has a big impact in the
synthesis process. Take, for instance, the ADT defined as $\textrm{data}\ T =
C_1\ T$, the synth goal $T \lolli C$, and its derivation:
%
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}T \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of}\ C_1\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of}\ C_1\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of}\ C_1\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}T \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
    T \lolli C \Uparrow}
\end{mathpar}
%
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Symmetrically, the
derivation for the synth goal $T$ is also infinite.
%
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
\end{mathpar}
%
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and, simetrically, don't allow construction of an
ADT when trying to synth an argument for its constructor. \todo{R: A segunda parte confusa era da frase de
  cima?, está melhor?}
%
%
For this, we need two more contexts, $\Rho_C$
for constraints on construction and $\Rho_D$ for constraints on
deconstruction. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
%
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    T \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        T \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    T\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}
%
These modifications prevent the infinite derivations in the scenarios described
above, however, they also greatly limit the space of derivable programs, leaving
the synthetiser effectively unable to synth from specifications with recursive
types\todo{refrasear tendo em conta que se sintetiza a partir de specs e nao
specs R: refrasear mas só com "from" não fica bem?}, which is so undesirable that we add three rules to complement the
restrictions on construction and destruction of recursive types.

@@

\mypara{Recursion} The main idea behind synthesis of recursive programs is the
labeling of the main goal and the addition of its type, under that name, to the
unrestricted context. That is, to synth a function of type $A \lolli B$ named
\emph{f}, the initial judgment can be written as
\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, f{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($f{:}A \lolli B$) in the $\Gamma$ context too.
We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.

@@

\todo[inline]{Polymorphic R: ??}

\mypara{Polymorphic Types} A polymorphic specification is a type of form
$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
bound variables that quantify universally the type variables in $\tau$, which is
also called a \emph{scheme}.
Synthesis of a scheme comprises of turning it into a non-quantified type, and
then treating its type variables agnostically. \todo{R: Melhor que
straightforward?}
First, \textbf{type variables are considered atomic types}, then, instantiate
the bound variables of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha
\lolli \alpha$ could be instantiated to
$\alpha0 \lolli \alpha0$. We add a rule for this:
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
where
\[
    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
    \overline{\alpha}.\ \tau
\]
%
As such, the construction of a derivation in which, to some types, no specific
rules can be applied (except for the \textsc{init} rule, a type variable can
instance a goal that's a type variable of the same name), corresponds to the
synthesis of a program where some expressions are treated agnostically --
nothing constrains their type -- meaning the program is polymorphic. The
simplest example is the polymorphic function \emph{id} of type $\forall \alpha
.\ \alpha \lolli \alpha$. The
program synthed from that specification is $\lambda x \rightarrow x$, a lambda
abstraction that doesn't constrain its parameter $x$ in any way. \todo{R :
rewrite this? não estranho?}

>>>>

\usepackage{minted}

@@

% \usemintedstyle{lovelace}
\usemintedstyle{murphy}

\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita - Nº 55902 \\
    Supervisor: Bernardo Toninho}
\date{July 2021}
\institute{NOVA School of Science and Technology}

@@

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program (i.e.~generating code) from a high-level specification.
%
Synthesis can be seen as a means to improve a programmer's productivity and/or
program correctness (e.g. through suggestion and/or autocompletion), or as a
tool to automate certain parts of the programming process, to name a few
examples.
% TODO: Escrever um início melhor

Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind of
specification, program synthesizers have to deal with two main inherent sources
of complexity: searching over a vast space of (potentially) valid programs, and
interpreting user intent.
%
In this work we'll explore type-based synthesis where specifications take the
form of linear types, which limit resource usage in programs, in combination
with polymorphism and refinements.  Concretely, we will explore the proof-theoretic foundations
of linear types via the Curry-Howard correspondence with linear logic, viewing
program synthesis under the lens of (linear logic) bottom-up proof search.

@@

At the core of the synthesis process, and of the developed synthetiser \synname,
is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
resource management. It connects the concepts seen in
background~\ref{sec:background} -- a type system
consisting of inference rules in sequent calculus for linear logic connectives;
with resource management, and focusing, as formalized techniques to,
respectively, solve resource related non-determinism, and to eliminate from the
proof search don't-care non-determinism.  Guided solely by this system's formal
representation\ref{sec:formal_system}, we can (and did) create a very robust
synthetiser for a select part of the linear types that's able to synthetise
programs successfully from e.g. the major linear logic equivalences. For example,
given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
the program $\lambda a .\ \lambda b .$ let $c \tensor d$ =
$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.

@@

However, a programming language only with the canonical linear types lacks
expressiveness. To enhance \synname\ and its language, synthetise more interesting
programs -- and thus empirically proving the feasability of more relevant
synthesis in a linear context, we extend the syntax and type system with
recursiveness and richer
types: -- polimorfic, refinements, adts parametrizáveis...  todo

\section{Synthesis Framework}\label{sec:formal_system}
% with Linear Types, Polymorphism and Refinements

% \todo[inline]{Encurtar titulo, mas algo nesta onda que descreve
%   ``tudo''}

@@

With this section we aim to formalize the rules that guide
the synthesis process for more expressive specifications, alongside the already
well defined rules that model the core of the synthetiser, resulting in
a \emph{sound} set of rules that characterizes our framework for synthesis of
recursive functions from polymorphic and refinement types in a linear context
% TODO: Better?....
, and describes the system in enough detail for
the synthetiser to be reproducible by a theory-driven implementation.

Even so, note that \emph{soundness} and \emph{completeness} have two very
different meanings, and our \emph{sound} set of rules only guarantees we won't
synthetise incorrect programs, but still hides a possibly infinite number of correct
proofs; and that the valid programs derivable through it reflect the somewhat
subjective trade-offs we decided on. Different choices and approaches outside
the core might lead to completely distinct synths and spaces of valid programs.

\mypara{Core Rules} The core system comprises of the sequent calculus rules for
linear logic connectives with resource-management and focusing.
\todo{Introdução para focusing...}
Focusing separates the proof search into two
phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later try to
invert, moving them to the linear context ($\Delta$) when we fail to.

@@

Putting together linear logic, the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synthetise a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

    % -o R
    \infer*[right=($\lolli R$)]
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
    \notin \Delta'}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x . M : A
    \lolli B \Uparrow}

@@

The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synthetise recursive
programs from more expressive (richer) types.

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a named sum type of any type, i.e. a named type that can be instantiated by one of many
tags (also known as constructors) that take exactly one fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which might be constructed with zero, one, or two
propositions of type $A$, using its syntax: \mintinline{haskell}{data Container
= None 1 | One A | Two (A * A)}

@@

@@

where the ADT $T$ and its constructors stand for any ADT with form
\begin{minted}{haskell}
    data T = C1 X1 | C2 X2 | ... | Cn Xn
\end{minted}

@@

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning constructors can take as arguments
values of the type they are defining. Seemingly small, this change has a big
impact in the synthesis process. Take, for instance, the ADT defined as
\mintinline{haskell}{data T = C1 T}, the synth goal $T \lolli C$, and its
derivation:
%
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}T \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of}\ C_1\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of}\ C_1\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of}\ C_1\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}T \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x . \dots :
    T \lolli C \Uparrow}
\end{mathpar}
%
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Symmetrically, the
derivation for the synth goal $T$ is also infinite.
%
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
\end{mathpar}
%
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and, symetrically, don't allow construction of an
ADT when trying to synthetise an argument for its constructor.
%
%
For this, we need two more contexts, $\Rho_C$
for constraints on construction and $\Rho_D$ for constraints on
deconstruction. Together, they hold a list of ADTs that can not be constructed or
deconstructed at some point in the proof. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
%
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    T \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        T \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    T\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}
%
These modifications prevent the infinite derivations in the scenarios described
above, however, they also greatly limit the space of derivable programs, leaving
the synthetiser effectively unable to synthetise from specifications with recursive
types, which is undesirable, so we add three rules to complement the
restrictions on construction and destruction of recursive types.

@@

\mypara{Recursion} The main idea behind synthesis of recursive programs is the
labeling of the main goal and the addition of its type, under that name, to the
unrestricted context. That is, to synthetise a function of type $A \lolli B$ named
\emph{f}, the initial judgment can be written as
\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, f{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($f{:}A \lolli B$) in the $\Gamma$ context too.
We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.

@@

\mypara{Polymorphic Types} A polymorphic specification is a type of form
$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of bound
variables that quantify universally the type variables in $\tau$, which is also
called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into a
non-quantified type, and then treating its type variables agnostically.  First,
\textbf{type variables are considered atomic types}, then, instantiate the bound
variables of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
this:
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
where
\[
    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
    \overline{\alpha}.\ \tau
\]
%
As such, the construction of a derivation in which, to some types, no specific
rules can be applied (except for the \textsc{init} rule, a type variable can
instance a goal that's a type variable of the same name), corresponds to the
synthesis of a program where some expressions are treated agnostically --
nothing constrains their type -- meaning the program is polymorphic. The
simplest example is the polymorphic function \emph{id} of type $\forall \alpha
.\ \alpha \lolli \alpha$. The
program synthetised from that specification is $\lambda x . x$, a lambda
abstraction that doesn't constrain its parameter $x$ in any way.
% TODO: \todo{R : rewrite this? não estranho?}
